%%% Text from the "idealised alt" section, common to both the TR and the
%%% submitted version. 

We can then compare the CSP model of an alt implementation to the idealised
model.  We can verify that the implementation model refines |AltSpec|$_F$ and
|AltSpec|$_{FD}$ in the relevant semantic models.

\begin{window}[1,r,{
%
\vspace{0.5ex}
\begin{tabular}{\|cccc\|}
Model & Threads & States  & Time\\
F & 4 & 33.1M & 21s \\
FD & 4 & 5.6M & 5.4s \\
F & 5 & 601M & 472s \\
FD & 5 & 62.7M & 46s \\
F & 6 & 10.6B & 153min \\
FD & 6 & 662M & 701s \\
\end{tabular}
},]
%
The table to the right gives statistics about these checks.  In each case, we
considered an alt with two branches, one in-port branch and one out-port
branch.  The checks are faster than the corresponding checks for a channel,
mainly because normalisation of the specification was
faster than for a channel, because the idealised alt has fewer states than an
idealised channel.  We think the main reason for this is that the |Lin(t)|
processes in the idealised alt have fewer states than the |Lin(t)| processes
in the idealised channel (there is one such process for each channel thread,
so the total number of states increases exponentially).  In addition, the
implementation model for an alt has fewer states than for a channel.
\end{window}

We used a couple of techniques to improve the efficiency of these checks.
Firstly, we applied the prioritisation operator of FDR to |IdealisedAlt| to
give priority to error events over all other events (except $\tau$s); so in
any state where an error event (or $\tau$) is possible, all other events are
blocked.  This is sound here because the subsequent behaviours allowed by the
specification (in |CHAOS(Interface)| or |DIV|) include the behaviours
corresponding to the blocked events.

Further, for the check in the stable-failures model, we normalised
|IdealisedAlt| before the application of the throws operator.  This made the
subsequent normalisation of |AltSpec|$_F$ faster.  (However, the same
technique in the failures-divergences model made the check slower.)

