\documentclass[12pt,a4paper]{article}

\usepackage{tikz}
\usetikzlibrary{arrows}
\def\X{node {$\cross$}}
\def\inCircle#1{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt}{#1}}}}
\def\m{$\mathord{\mid}$}

\usepackage{scalalistings}
\unScalaMid  % We need "|" to have standard meaning when cspm.sty is read
\usepackage{cspm}

\def\cspmsize{\footnotesize}
\def\cspmDisplaySize{\footnotesize}
%\def\cspmfont{\itshape}
% Macros to control whether |...| produces Scala or CSP formatting.
\def\inlineScala{\uncspMid\scalaMid}
\def\inlineCSP{\unScalaMid\cspMid}
%\inlineCSP
\cspMid

\def\topfraction{0.8}
\renewcommand{\floatpagefraction}{0.7}
\sloppy

\title{Analysing a Library of Concurrency Primitives using CSP}
\author{Gavin Lowe}

\begin{document}
\maketitle

\section{Introduction}

Overview of SCL.  Channels, alts, closing, timeouts.  Guards.  Restrictions.
Continuations. 

Main contributions to emphasise:
%
\begin{itemize}
\item Specification of synchronisation.

\item Difficulty of capturing correct behaviour when the channel is closed. 

\item Direct versus compositional verification. 

\end{itemize}

One-page summary of modelling technique.  Variables, monitors.  Assertions.
Continuation-passing? 

Overview of basic model.  |beginSend|, |endSend| events.

Specification of this, capturing synchronisations.

Adding closing.  Extending specification.  Why the previous implementation
went wrong. 

Adding timed operations.

Overview of alt protocol.  Locking. 

Extension of channel model:  Registering and deregistering.  call backs:
|beginMaybeReceive|, |endMaybeReceive|, etc. 

Model of alt.  Interface. 

Direct analysis.  Specification.  Limitations.

Compositional verification idea.

Idealised channel; analysis.

Idealised alt; analysis.

Compositional verification. 

%%%%%%%%%%

%% \section{Building blocks}

%% Variables, continuation passing, computations, JVM monitors, SCL monitors. 

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Overview of basic model.  |beginSend|, |endSend| events.

\input{syncchan} % Basic model of SyncChan

\input{syncchan2} % Closing channels

\input{alt} % Introduction to alts

\input{combinedTest}

\input{compositionalTest}

\input{idealisedChan}

\input{idealisedAlt}



\end{document}
 
